$Z_3$ topological order in the quantum dimer-pentamer model

### 2017-07-07 qdpm_paper.tex
Removed unecessary equation tags
Removed comments for arXiv
Added citation to trimer paper and submission letter
Worked on referee response

%\newcommand*\bplqt{\includegraphics[height=1.6ex] {Symbols/blank_plqt}}
%\newcommand*\bhrzmv{\includegraphics[height=1.6ex]{Symbols/blank_hrzmv}}
%\newcommand*\bdiag{\includegraphics[height=1.6ex] {Symbols/blank_diag}}
%\newcommand*\bstr{\includegraphics[height=1.6ex] {Symbols/blank_str}}

@@

%\newcommand*{\hprs}{%
%  \text{% change size in subscripts or superscripts
%    \includegraphics[
%      height=1.2ex,% adjust to suit
%      valign=M,% center vertically
%      raise=\fontdimen22\textfont2,% but raise it to the formula axis
%    ]{Symbols/hrzprestar} }
%}
%\newcommand*{\hspr}{%
%  \text{% change size in subscripts or superscripts
%    \includegraphics[
%      height=1.2ex,% adjust to suit
%      valign=M,% center vertically
%      raise=\fontdimen22\textfont2,% but raise it to the formula axis
%    ]{Symbols/hrzstarpair} }
%}

@@

%\newcommand*{\bhrzmv}{%
%  \text{% change size in subscripts or superscripts
%    \includegraphics[
%      height=1.2ex,% adjust to suit
%      valign=M,% center vertically
%      raise=\fontdimen22\textfont2,% but raise it to the formula axis
%    ]{Symbols/blank_hrzmv} }
%}
%\newcommand*{\bdiag}{%
%  \text{% change size in subscripts or superscripts
%    \includegraphics[
%      height=1.2ex,% adjust to suit
%      valign=M,% center vertically
%      raise=\fontdimen22\textfont2,% but raise it to the formula axis
%    ]{Symbols/blank_diag} }
%}
%\newcommand*{\bstr}{%
%  \text{% change size in subscripts or superscripts
%    \includegraphics[
%      height=1.2ex,% adjust to suit
%      valign=M,% center vertically
%      raise=\fontdimen22\textfont2,% but raise it to the formula axis
%    ]{Symbols/blank_str} }
%}

@@

%\newcommand*{\threedownplqt}{%
%  \text{% change size in subscripts or superscripts
%    \includegraphics[
%      height=1.5ex,% adjust to suit
%      valign=M,% center vertically
%      raise=\fontdimen22\textfont2,% but raise it to the formula axis
%    ]{Symbols/threedownplqt.pdf} }
%}
%\newcommand*{\threeupplqt}{%
%  \text{% change size in subscripts or superscripts
%    \includegraphics[
%      height=1.5ex,% adjust to suit
%      valign=M,% center vertically
%      raise=\fontdimen22\textfont2,% but raise it to the formula axis
%    ]{Symbols/threeupplqt.pdf} }
%}
%\newcommand*{\threerightplqt}{%
%  \text{% change size in subscripts or superscripts
%    \includegraphics[
%      height=1.5ex,% adjust to suit
%      valign=M,% center vertically
%      raise=\fontdimen22\textfont2,% but raise it to the formula axis
%    ]{Symbols/threerightplqt.pdf} }
%}

@@

%\newcommand*{\onedownplqt}{%
%  \text{% change size in subscripts or superscripts
%    \includegraphics[
%      height=1.5ex,% adjust to suit
%      valign=M,% center vertically
%      raise=\fontdimen22\textfont2,% but raise it to the formula axis
%    ]{Symbols/onedownplqt.pdf} }
%}
%\newcommand*{\oneupplqt}{%
%  \text{% change size in subscripts or superscripts
%    \includegraphics[
%      height=1.5ex,% adjust to suit
%      valign=M,% center vertically
%      raise=\fontdimen22\textfont2,% but raise it to the formula axis
%    ]{Symbols/oneupplqt.pdf} }
%}
%\newcommand*{\onerightplqt}{%
%  \text{% change size in subscripts or superscripts
%    \includegraphics[
%      height=1.5ex,% adjust to suit
%      valign=M,% center vertically
%      raise=\fontdimen22\textfont2,% but raise it to the formula axis
%    ]{Symbols/onerightplqt.pdf} }
%}

@@

Strongly interacting quantum many-body systems are known to support a variety of exotic quantum phases of matter with no classical analogue. Among these are topologically ordered quantum liquids; their non-trivial topological order can not be identified by a local order parameter as these phases of matter don't break conventional symmetries~\cite{Wen1990,Nayak2008}. Both because of of the exotic nature of topological phases and their potential use for quantum information processing~\cite{Freedman2001,Kitaev2003}, there is great interest in the development of theoretical models which display these phases and could provide candidate experimental systems to constructively engineer topological order~\cite{Duan2003,Jaksch2005,Lewenstein2007,Jiang2008d,Weimer2010a,Herdman2010c,Fowler2012a}.

@@

While on non-bipartite two-dimensionsal lattices, the QDM has a topologically ordered ground state~\cite{Moessner2001a,Fendley2002}, on the square lattice there is only a symmetric liquid ground state at an isolated critical quantum point~\cite{Leung1996,Syljuasen2006}. If the hard-core constraint of the square lattice QDM is relaxed such that only the parity of the number of dimer touching each vertex is fixed, the simplest non-trivial dynamics leads to a $Z_2$ topologically ordered ground state~\cite{Kitaev2003,Wen2003}. The existence of topologically ordered ground states in these models is intimately related to the local constraint which imposes a local gauge symmetry in the Hilbert space; the topological order that arises is directly related to a discrete gauge theory with the appropriate local gauge constraint~\cite{Moessner2001}.

Given that these models with dimer degrees of freedom and local constraints posses the simplest non-trivial local Hilbert space at each link, they provide a potentially simpler route to engineering exotic phases. The local constraints can be enforced by a local potential energy penalty and the minimal dynamics often arise at the lowest non-trivial order in perturbation theory from a related spin model\cite{Albuquerque2008}. We will refer to such models with dimer degrees of freedom on links and local constraints as \emph{generalized quantum dimer modes} (GQDMs). Previous work has demonstrated that GQDMs on the square lattice support gapped $Z_2$ topologically ordered phases as well as gapless quantum critical points. On non-bipartite lattices, GQDMs support $Z_2$ topological order~\cite{Moessner2001a,Misguich2002} as well as doubled-semion phases~\cite{Qi2015,Buerschaper2014}. A full characterization of all exotic phases that can exist within this framework is thus desirable.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[htpb]
    \centering
    \includegraphics[width=0.9\linewidth]{example_local_constraints.pdf}
    \caption{Examples of GQDMs with different local constraints. Clockwise from the upper left panel; fixed dimer number: $n_d=1$, $n_d=2$;  fixed dimer parity: odd, even.}
    \label{fig:example_local_constraints}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Here we will provide background about generalized quantum dimer models on the square lattice. We consider systems with Ising-like dimer degrees of freedom on each link, such that each link is either unoccupied or occupied by a single dimer. Given an Ising spin $\sigma_\ell$ on each link $\ell$ we define the number of dimers on the link $n_\ell \equiv 1/2+\sigma_\ell^z$, where $\sigma_\ell^i$ are the associated Pauli matrices. The Hilbert space is spanned by an orthonormal basis of dimerizations of the square lattice that satisfy a local constraint at each vertex. We define a $U(1)$ charge $Q_v$ at each vertex $v$ of the lattice in terms of the number of dimers touching $v$, $n_v\equiv \sum_{\ell \in v}n_\ell$:
\begin{equation}
Q_v \equiv \pm \left( n_v - c_P \right)
\end{equation}
where $c_P = 2$, ($c_P=1$) for models in the even (odd) sector of the constraint, and the $+$($-$) sign corresponds to vertices on the $A$($B$) sublattice. A local dimer number constraint requires $Q_v$ to vanish for all vertices; a dimer parity constraint requires $(Q_v\mod 2)$ to vanish.

The Hamiltonian of a GQDM is of the form
\begin{equation}
H_{\rm{GQDM}} = -\sum_\nu t_\nu  {T}_{\nu} + \sum_\nu v_\nu  {V}_{\nu} \label{eq:HGQDM}
\end{equation}
where ${T}_\nu$ are off-diagonal kinetic energy operators, ${V}_\nu$ are diagonal potential energy operators, $t_\nu$ and $v_\nu$ parametrize the strength of these terms, and each $\nu$ represents a set of  neighboring links. We will limit our discussion to $t_\nu >0$. ${T}_\nu$ is constructed to generate resonances between two distinct allowed dimerazations of $\nu$, $c_\nu$ and its compliment $\bar{c}_\nu$, and is written as:
\begin{equation}
 {T}_{\nu} = \eket{c_\nu}\ebra{\bar{c}_\nu} + h.c.;
\end{equation}
in the simplest case, $ {T_\nu}$ involves the minimum non-trivial dimer dynamics that preserve the constraint. Similarly, the potential energy term provides an energy cost for states with these local configurations:
\begin{equation}
 {V}_{\nu} = \eket{c_\nu}\ebra{c_\nu} +  \eket{\bar{c}_\nu}\ebra{\bar{c}_\nu}.
\end{equation}
Therefore we may define a Hamiltonian for a GQDM using \Eqref{eq:HGQDM} and defining a set of dimer resonances $\{(c_\nu,\bar{c}_\nu)\}$ and their corresponding strengths $\{(t_\nu,v_\nu)\}$:
\begin{equation}
\HGQDM \biggl( \bigl\{\left(c_\nu,\bar{c}_\nu\right)\},\{\left(t_\nu,v_\nu\right)\bigr\}\biggr).
\end{equation}
 In general, $[T_\nu, T_{\nu'}] \neq 0 $ and $[T_\nu,V_{\nu'}]\neq0$ when $\nu$ and $\nu'$ share links, and thus the ground state of $H_{\rm{GQDM}}$ is not explicitly known.

However, at the special points in the $\{(t_\nu,v_\nu)\}$ parameter space, known as the Rokhsar-Kivelson (RK) points, when \Eqref{eq:HGQDM} can be written as the sum of projection operators
\begin{equation}
H_{\rm{RK}} = \sum_\nu  {h}_\nu \label{eq:HRK},
\end{equation}
where all $h_\nu$ are projectors with eigenvalues $\{0,1\}$, the exact ground state is known to be
the zero energy state of \Eqref{eq:HRK} that is simultaneously annihilated by all ${h}_\nu$. Such a zero energy state, known as the RK state, may be formed from an equal superposition of all dimerizations that are connected by $\{T_\nu\}$:
\begin{equation}
\eket{\Phi_{\RK}^{\Omega}} = \sum_{C \in \Omega} \eket{C}
\end{equation} 
where $\Omega$ is a subset of the Hilbert space connected by $\{T_\nu\}$. Thus $\{ \ket{\Phi_{\RK}^\Omega} \}$ are the exact, degenerate ground states of $H_{\rm{GQDM}}$ with a degeneracy labelled by $\Omega$ at the RK point.

@@

On surfaces with non-trivial topology, we may define the topological loop operators
\begin{equation}
W_{X} \left( \alpha \right) \equiv S_{\Gamma_{X}} \left( \alpha \right)
\end{equation}
where $\Gamma_{X}$ is a topologically non-trivial closed loop that winds once about the $x$ axis; we define $W_Y$ similarly. Since $W_{X} (\alpha)$ and $W_{Y} (\alpha)$ commute with the Hamiltonian, we may use their eigenvalues to label the ground states.

@@

The Hamiltonian for these models gives dynamics to parallel dimers on the plaquettes of the lattice:
\begin{equation}
\HQDM\left(t,v\right) \equiv \HGQDM \biggl( \bigl\{ \left(\vrtplqt,\hrzplqt\right) \bigr\},\bigl\{ \left(t,v\right) \bigr\} \biggr)  \label{eq:HQDM}
\end{equation}
where we have chosen $t_\bplqt = t$, $v_\bplqt =v, \forall \bplqt$.
%\begin{align}
%    \label{}
%   \HQDM = &-t\sum_{\bplqt} \biggl( \eket{\vrtplqt} \ebra{\hrzplqt} +h.c. \biggr) \notag\\
%    					&+v\sum_{\bplqt} \biggl( \eket{\vrtplqt} \ebra{\vrtplqt} +\eket{\hrzplqt} \ebra{\hrzplqt} \biggr). \label{eq:HQDM}
%\end{align}
In this case of a fixed number constraint, the Hamiltonian conserves $Q_v$ and therefore the dimer flux through any closed loop such that:
\begin{equation}
\left[ \Phi_\Gamma, \HQDM \right] = 0.
\end{equation}
We can define two conserved topological winding numbers $\Phi_{X}$ \& $\Phi_{Y}$ which are the dimer flux through topologically non-trivial loops that wind around each axis of the lattice when the system has periodic boundary conditions (see \figref{fig:orientedSL}). On a lattice of dimensions $L_x\times L_y$, the eigenvalues of $\Phi_X$ and $\Phi_Y$ are integers in the range
\begin{equation}
-L_{x}/2 \leq  \Phi_{X} \leq L_{x}/2,
\end{equation}
and similarly for $\Phi_Y$, thus there are $\mathcal{O} (L_x L_y)$ distinct topological sectors.

At the Rokhsar-Kivelson point $t=v$, we may write $\HQDM$ in the form of \Eqref{eq:HRK} by defining projection operators  $ {h}_\nu $:
\begin{equation}
 {h}_\nu \equiv \biggl( \eket{ c_\nu} - \eket{\bar{c}_\nu} \biggr)\biggl( \ebra{ c_\nu} - \ebra{\bar{c}_\nu} \biggr).
\end{equation} 
Since the plaquette flip term in \Eqref{eq:HQDM} is believed to be ergodic in each of topological sector $(\Phi_x,\Phi_y)$ on the square lattice, each topological sector corresponds to a unique RK ground state where $\Omega$ is labeled by $(\Phi_x,\Phi_y)$. Thus the  RK points of the QDM and QFPLM have an \emph{extensively} degenerate ground states.

@@

We now relax the local constraint to a parity constraint such that only $Q_v^{Z_2} \equiv (Q_v \mod 2)$ is required to vanish; the resulting GQDM are the even and odd sectors are the  even \& odd Ising gauge theories (IGT) with constraints $n_v \in \{0,2,4\}$ and $n_v \in \{1,3\}$, respectively. In these cases, the local constraint requires $\alpha_v \in \{ 0, \pi\}$ in \Eqref{eq:Gv} and therefore the gauge symmetry is reduced from $U(1)$ to $Z_2$. The set of allowed dimer resonances $\mathcal{C}_{\rm{IGT}}$ is expanded accordingly:
\begin{align}
\mathcal{C}_{\rm{IGT}} &\equiv \bigl\{ \left(c_{\Box,i},\bar{c}_{\Box,i}\right) \bigr\} \\
&=\bigl\{
\left(\vrtplqt,\hrzplqt\right),\left(\threeleftplqt,\oneleftplqt\right),\left(\twodownplqt,\twoupplqt\right),\left(\bplqt,\fplqt\right) \bigr\}
\end{align}
where we implicitly include symmetry related resonances. We will limit our discussion to $v_\bplqt=0$ such that the corresponding Hamiltonian is
\begin{equation}
\HIGT\left(t\right) \equiv \HGQDM \biggl( \mathcal{C}_{\rm{IGT}},\bigl\{ \left(t,0\right) \bigr\} \biggr).
\end{equation}
In this case the equal superposition RK wavefunction is the exact ground state and occurs in the absence of a potential energy term, as we can define a projection operator on each plaquette
\begin{equation}
h_\Box \equiv \frac{1}{2} \left( \mathbb{1} - \sum_i \eket{c_{\Box,i}}\ebra{\bar{c}_{\Box,i}} + h.c. \right)
\end{equation}
such that $\HIGT$ takes the form of \Eqref{eq:HRK} (up to a constant), and the RK state is the ground state for all values of $t$.

@@

We define the Hilbert space of the QDPM to be a GQDM where vertices are constrained to be touched by a single dimer or the center of a pentamer, such that $n_v \in \{1,4\}$. For defect-free configurations in the QDPM, $Q_v$ only vanishes modulo $3$;  thus we may define a $Z_3$ charge,
\begin{equation}
Q_v^{Z_3} \equiv \left(Q_v~\rm{mod}~3\right)
\end{equation}
 which strictly vanishes at each vertex, representing the local constraint. All states in the QDPM are invariant under an exact local $Z_3$ gauge symmetry given by \Eqref{eq:Gv} with $\alpha_v \in \{0,\pm 2\pi/3\}$. Thus introducing pentamers in the QDM reduces the gauge symmetry from $U(1)$ to $Z_3$.

The Hamiltonian of the QDPM is defined by \Eqref{eq:HGQDM} including both plaquette flips as well as local dimer resonances that create and destroy pairs of pentamers, and allow single pentamers to translate on the same sub-lattice; these local dynamics involving pentamers are illustrated in  \tabref{tab:pentamer_moves}. Note that pentamers give dynamics to ``staggered" regions with no flippable plaquettes; thus we expect the QDPM to have weaker correlations that the QDM, where the dynamics are more constrained. We may write $\HQDPM$ as the following:
\begin{equation}
\HQDPM \equiv \HQDM + H_{\rm{pen}}
\end{equation}
where $H_{\rm{pen}}$ only involves the pentamer resonances in \tabref{tab:pentamer_moves} and those related by symmetry:
\begin{align}
H_{\rm{pen}} \equiv \HGQDM \biggl( &\bigl\{ \left(c_{\nu_1},\bar{c}_{\nu_1}\right),\left(c_{\nu_2},\bar{c}_{\nu_2}\right),\left(c_{\nu_3},\bar{c}_{\nu_3}\right) \bigr\},\biggr.\notag\\
\biggl. &\bigl\{ \left(t_1,v_1\right),\left(t_2,v_2\right),\left(t_3,v_3\right) \bigr\} \biggr)  \label{eq:Hpen}
\end{align}

@@

Our numerical simulations (discussed below in \secref{sec:numerics}) suggest that the dynamics of $\HQDPM$ are ergodic in each topological sector $\Omega$ that are labeled by the topological loop operators:
\begin{equation}
\Omega = \left(W_X^{Z_3},W_Y^{Z_3} \right).
\end{equation}
The exact zero energy ground state at the RK point, where $t_i=v_i, i \in \{ 1,2,3\}$, is the RK state, where the superposition is taken to be over all states in the topological sector. Since the $W_{X}^{Z_3}$ and  $W_{Y}^{Z_3}$ have eigenvalues $\{1,e^{i2\pi/3},e^{-i2\pi/3}\}$, the RK point of the QDPM is therefore 9-fold degenerate on the torus.

@@

Additionally we have studied the nature of the decay of the connected dimer-dimer correlation function,
  \begin{equation}
    C_{\mathrm{d}} \left(r\right) \equiv \langle n_{\bm{0}} n_{\bm{r}} \rangle - \langle n_{\bm{0}} \rangle   \langle n_{\bm{r}} \rangle   
\end{equation}
where $r$ is the distance between the links $\bm{0}$ and $\bm{r}$. In \figref{fig:spatial_dmr_cor}, $C_d$ between parallel links is plotted as a function of separation, for $\bm{r}$ perpendicular to the orientation of the dimers for a $L=64$ lattice. We find that $C_d$ decays exponentially with distance having a correlation length  $\chi\approx 1$. Additionally, $C_d$ displays a sub-lattice oscillation that is consistent with the finite-size maximum in $S_d$ at $(0,\pi)$. Thus we find that the RK point of the QDPM is a symmetric dimer liquid with exponentially decaying dimer correlations, in contrast with the power-law decaying correlations of the QDM.

@@

\section{Dimer order parameters QDPM RK ground state}
\label{sec:dimer_order_parameters}
%
%%%%%%%%%%%%%%%%%%%%%%%%
%\begin{figure}
%    \centering
%    \includegraphics[width=1.0\columnwidth]{spatial_cors_end_to_end_parallel.pdf}
%    \caption{Correlation function between parallel dimers (end-to-end) on the same (blue) and different (red) 
%    sub-lattice for a $L=64$ lattice.}
%    \label{fig:spatial_dmr_cor}
%\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

@@

@@

@@

@@

@@

Strongly interacting quantum many-body systems are known to support a variety of exotic quantum phases of matter with no classical analogue. Among these are topologically ordered quantum liquids; their non-trivial topological order can not be identified by a local order parameter as these phases of matter don't break conventional symmetries~\cite{Wen1990,Nayak2008}. Both because of the exotic nature of topological phases and their potential use for quantum information processing~\cite{Freedman2001,Kitaev2003}, there is great interest in the development of theoretical models which display these phases and could provide candidate experimental systems to constructively engineer topological order~\cite{Duan2003,Jaksch2005,Lewenstein2007,Jiang2008d,Weimer2010a,Herdman2010c,Fowler2012a}.

@@

While on non-bipartite two-dimensionsal lattices, the QDM has a topologically ordered ground state~\cite{Moessner2001a,Fendley2002}, on the square lattice there is only a symmetric liquid ground state at an isolated critical quantum point~\cite{Leung1996,Syljuasen2006}. If the hard-core constraint of the square lattice QDM is relaxed such that only the parity of the number of dimers touching each vertex is fixed, the simplest non-trivial dynamics leads to a $Z_2$ topologically ordered ground state~\cite{Kitaev2003,Wen2003}. The existence of topologically ordered ground states in these models is intimately related to the local constraint which imposes a local gauge symmetry in the Hilbert space; the topological order that arises is directly related to a discrete gauge theory with the appropriate local gauge constraint~\cite{Moessner2001}.

Given that these models with dimer degrees of freedom and local constraints posses the simplest non-trivial local Hilbert space at each link, they provide a potentially simpler route to engineering exotic phases. The local constraints can be enforced by a local potential energy penalty and the minimal dynamics often arise at the lowest non-trivial order in perturbation theory from a related spin model\cite{Albuquerque2008}. We will refer to such models with dimer degrees of freedom on links and local constraints as \emph{generalized quantum dimer modes} (GQDMs). Previous work has demonstrated that GQDMs on the square lattice support gapped $Z_2$ topologically ordered phases as well as gapless quantum critical points. On non-bipartite lattices, GQDMs support $Z_2$ topological order~\cite{Moessner2001a,Misguich2002} as well as doubled-semion phases~\cite{Qi2015,Buerschaper2014}. Additionally, recent work on a trimer resonating valence bond state suggests that it may be described by a $Z_3$ gauge theory~\cite{Lee2017}. A full characterization of all exotic phases that can exist within this framework is thus desirable.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[htpb]
    \centering
    \includegraphics[width=0.9\linewidth]{example_local_constraints.pdf}
    \caption{Examples of GQDMs with different local constraints. Clockwise from the upper left panel; fixed dimer number: $n_d=1$, $n_d=2$;  fixed dimer parity: odd, even.}
    \label{fig:example_local_constraints}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Here we will provide background about generalized quantum dimer models on the square lattice. We consider systems with Ising-like dimer degrees of freedom on each link, such that each link is either unoccupied or occupied by a single dimer. Given an Ising spin $\sigma_\ell$ on each link $\ell$ we define the number of dimers on the link $n_\ell \equiv 1/2(1+\sigma_\ell^z)$, where $\sigma_\ell^i$ are the associated Pauli matrices. The Hilbert space is spanned by an orthonormal basis of dimerizations of the square lattice that satisfy a local constraint at each vertex. We define a $U(1)$ charge $Q_v$ at each vertex $v$ of the lattice in terms of the number of dimers touching $v$, $n_v\equiv \sum_{\ell \in v}n_\ell$:
\begin{equation}
Q_v \equiv \pm \left( n_v - c_P \right) \notag
\end{equation}
where $c_p$ is a constant, and we define the even (odd) sector of the constraint to have $c_P = 2$ ($c_P=1$); the $+$($-$) sign corresponds to vertices on the $A$($B$) sublattice (see \tabref{tab:defects}). A local dimer number constraint requires $Q_v$ to vanish for all vertices; a dimer parity constraint requires $(Q_v\mod 2)$ to vanish (see \figref{fig:example_local_constraints}).

The Hamiltonian of a GQDM is of the form
\begin{equation}
H_{\rm{GQDM}} = -\sum_\nu t_\nu  {T}_{\nu} + \sum_\nu v_\nu  {V}_{\nu} \label{eq:HGQDM}
\end{equation}
where ${T}_\nu$ are off-diagonal kinetic energy operators, ${V}_\nu$ are diagonal potential energy operators, $t_\nu$ and $v_\nu$ parametrize the strength of these terms, and each $\nu$ represents a set of  neighboring links. We will limit our discussion to $t_\nu >0$. ${T}_\nu$ is constructed to generate resonances between two distinct allowed dimerazations of $\nu$, $c_\nu$ and its compliment $\bar{c}_\nu$, and is written as:
\begin{equation}
 {T}_{\nu} = \eket{c_\nu}\ebra{\bar{c}_\nu} + h.c.;  \notag
\end{equation}
in the simplest case, $ {T_\nu}$ involves the minimum non-trivial dimer dynamics that preserve the constraint. Similarly, the potential energy term provides an energy cost for states with these local configurations:
\begin{equation}
 {V}_{\nu} = \eket{c_\nu}\ebra{c_\nu} +  \eket{\bar{c}_\nu}\ebra{\bar{c}_\nu}.  \notag
\end{equation}
Therefore we may define a Hamiltonian for a GQDM using \Eqref{eq:HGQDM} and defining a set of dimer resonances $\{(c_\nu,\bar{c}_\nu)\}$ and their corresponding strengths $\{(t_\nu,v_\nu)\}$:
\begin{equation}
\HGQDM \biggl( \bigl\{\left(c_\nu,\bar{c}_\nu\right)\},\{\left(t_\nu,v_\nu\right)\bigr\}\biggr).  \notag
\end{equation}
 In general, $[T_\nu, T_{\nu'}] \neq 0 $ and $[T_\nu,V_{\nu'}]\neq0$ when $\nu$ and $\nu'$ share links, and thus the ground state of $H_{\rm{GQDM}}$ is not explicitly known.

However, at the special points in the $\{(t_\nu,v_\nu)\}$ parameter space, known as the Rokhsar-Kivelson (RK) points, when \Eqref{eq:HGQDM} can be written as the sum of projection operators
\begin{equation}
H_{\rm{RK}} = \sum_\nu  {h}_\nu \label{eq:HRK},
\end{equation}
where all $h_\nu$ are projectors with eigenvalues $\{0,1\}$, the exact ground state is known to be
the zero energy state of \Eqref{eq:HRK} that is simultaneously annihilated by all ${h}_\nu$. Such a zero energy state, known as the RK state, may be formed from an equal superposition of all dimerizations that are connected by $\{T_\nu\}$:
\begin{equation}
\eket{\Phi_{\RK}^{\Omega}} = \sum_{C \in \Omega} \eket{C}  \notag
\end{equation} 
where $\Omega$ is a subset of the Hilbert space connected by $\{T_\nu\}$. Thus $\{ \ket{\Phi_{\RK}^\Omega} \}$ are the exact, degenerate ground states of $H_{\rm{GQDM}}$ with a degeneracy labelled by $\Omega$ at the RK point.

@@

On surfaces with non-trivial topology, we may define the topological loop operators
\begin{equation}
W_{X} \left( \alpha \right) \equiv S_{\Gamma_{X}} \left( \alpha \right)  \notag
\end{equation}
where $\Gamma_{X}$ is a topologically non-trivial closed loop that winds once about the $x$ axis; we define $W_Y$ similarly. Since $W_{X} (\alpha)$ and $W_{Y} (\alpha)$ commute with the Hamiltonian, we may use their eigenvalues to label the ground states.

@@

The Hamiltonian for these models gives dynamics to parallel dimers on the plaquettes of the lattice:
\begin{equation}
\HQDM\left(t,v\right) \equiv \HGQDM \biggl( \bigl\{ \left(\vrtplqt,\hrzplqt\right) \bigr\},\bigl\{ \left(t,v\right) \bigr\} \biggr)  \label{eq:HQDM}
\end{equation}
where we have chosen $t_\bplqt = t$, $v_\bplqt =v, \forall \bplqt$. In this case of a fixed number constraint, the Hamiltonian conserves $Q_v$ and therefore the dimer flux through any closed loop such that:
\begin{equation}
\left[ \Phi_\Gamma, \HQDM \right] = 0.  \notag
\end{equation}
We can define two conserved topological winding numbers $\Phi_{X}$ \& $\Phi_{Y}$ which are the dimer flux through topologically non-trivial loops that wind around each axis of the lattice when the system has periodic boundary conditions (see \figref{fig:orientedSL}). On a lattice of dimensions $L_x\times L_y$, the eigenvalues of $\Phi_X$ and $\Phi_Y$ are integers in the range
\begin{equation}
-L_{x}/2 \leq  \Phi_{X} \leq L_{x}/2,  \notag
\end{equation}
and similarly for $\Phi_Y$, thus there are $\mathcal{O} (L_x L_y)$ distinct topological sectors.

At the Rokhsar-Kivelson point $t=v$, we may write $\HQDM$ in the form of \Eqref{eq:HRK} by defining projection operators  $ {h}_\nu $:
\begin{equation}
 {h}_\nu \equiv \biggl( \eket{ c_\nu} - \eket{\bar{c}_\nu} \biggr)\biggl( \ebra{ c_\nu} - \ebra{\bar{c}_\nu} \biggr).  \notag
\end{equation} 
Since the plaquette flip term in \Eqref{eq:HQDM} is believed to be ergodic in each of topological sector $(\Phi_x,\Phi_y)$ on the square lattice, each topological sector corresponds to a unique RK ground state where $\Omega$ is labeled by $(\Phi_x,\Phi_y)$. Thus the  RK points of the QDM and QFPLM have an \emph{extensively} degenerate ground states.

@@

We now relax the local constraint to a parity constraint such that only $Q_v^{Z_2} \equiv (Q_v \mod 2)$ is required to vanish; the resulting GQDM are the even and odd sectors are the  even \& odd Ising gauge theories (IGT) with constraints $n_v \in \{0,2,4\}$ and $n_v \in \{1,3\}$, respectively. In these cases, the local constraint requires $\alpha_v \in \{ 0, \pi\}$ in \Eqref{eq:Gv} and therefore the gauge symmetry is reduced from $U(1)$ to $Z_2$. The set of allowed dimer resonances $\mathcal{C}_{\rm{IGT}}$ is expanded accordingly:
\begin{align}
\mathcal{C}_{\rm{IGT}} &\equiv \bigl\{ \left(c_{\Box,i},\bar{c}_{\Box,i}\right) \bigr\} \\
&=\bigl\{
\left(\vrtplqt,\hrzplqt\right),\left(\threeleftplqt,\oneleftplqt\right),\left(\twodownplqt,\twoupplqt\right),\left(\bplqt,\fplqt\right) \bigr\}
\end{align}
where we implicitly include symmetry related resonances. We will limit our discussion to $v_\bplqt=0$ such that the corresponding Hamiltonian is
\begin{equation}
\HIGT\left(t\right) \equiv \HGQDM \biggl( \mathcal{C}_{\rm{IGT}},\bigl\{ \left(t,0\right) \bigr\} \biggr).  \notag
\end{equation}
In this case the equal superposition RK wavefunction is the exact ground state and occurs in the absence of a potential energy term, as we can define a projection operator on each plaquette
\begin{equation}
h_\Box \equiv \frac{1}{2} \left( \mathbb{1} - \sum_i \eket{c_{\Box,i}}\ebra{\bar{c}_{\Box,i}} + h.c. \right)  \notag
\end{equation}
such that $\HIGT$ takes the form of \Eqref{eq:HRK} (up to a constant), and the RK state is the ground state for all values of $t$.

@@

We define the Hilbert space of the QDPM to be a GQDM where vertices are constrained to be touched by a single dimer or the center of a pentamer, such that $n_v \in \{1,4\}$. For defect-free configurations in the QDPM, $Q_v$ only vanishes modulo $3$;  thus we may define a $Z_3$ charge,
\begin{equation}
Q_v^{Z_3} \equiv \left(Q_v~\rm{mod}~3\right)  \notag
\end{equation}
 which strictly vanishes at each vertex, representing the local constraint. All states in the QDPM are invariant under an exact local $Z_3$ gauge symmetry given by \Eqref{eq:Gv} with $\alpha_v \in \{0,\pm 2\pi/3\}$. Thus introducing pentamers in the QDM reduces the gauge symmetry from $U(1)$ to $Z_3$.

The Hamiltonian of the QDPM is defined by \Eqref{eq:HGQDM} including both plaquette flips as well as local dimer resonances that create and destroy pairs of pentamers, and allow single pentamers to translate on the same sub-lattice; these local dynamics involving pentamers are illustrated in  \tabref{tab:pentamer_moves}. Note that pentamers give dynamics to ``staggered" regions with no flippable plaquettes; thus we expect the QDPM to have weaker correlations than the QDM, where the dynamics are more constrained. We may write $\HQDPM$ as the following:
\begin{equation}
\HQDPM \equiv \HQDM + H_{\rm{pen}}  \notag
\end{equation}
where $H_{\rm{pen}}$ only involves the pentamer resonances in \tabref{tab:pentamer_moves} and those related by symmetry:
\begin{align}
H_{\rm{pen}} \equiv \HGQDM \biggl( &\bigl\{ \left(c_{\nu_1},\bar{c}_{\nu_1}\right),\left(c_{\nu_2},\bar{c}_{\nu_2}\right),\left(c_{\nu_3},\bar{c}_{\nu_3}\right) \bigr\},\biggr.\notag\\
\biggl. &\bigl\{ \left(t_1,v_1\right),\left(t_2,v_2\right),\left(t_3,v_3\right) \bigr\} \biggr)  \label{eq:Hpen}
\end{align}

@@

Our numerical simulations (discussed below in \secref{sec:numerics}) suggest that the dynamics of $\HQDPM$ are ergodic in each topological sector $\Omega$ that are labeled by the topological loop operators:
\begin{equation}
\Omega = \left(W_X^{Z_3},W_Y^{Z_3} \right).  \notag
\end{equation}
The exact zero energy ground state at the RK point, where $t_i=v_i, i \in \{ 1,2,3\}$, is the RK state, where the superposition is taken to be over all states in the topological sector. Since the $W_{X}^{Z_3}$ and  $W_{Y}^{Z_3}$ have eigenvalues $\{1,e^{i2\pi/3},e^{-i2\pi/3}\}$, the RK point of the QDPM is therefore 9-fold degenerate on the torus.

@@

Additionally we have studied the nature of the decay of the connected dimer-dimer correlation function,
  \begin{equation}
    C_{\mathrm{d}} \left(r\right) \equiv \langle n_{\bm{0}} n_{\bm{r}} \rangle - \langle n_{\bm{0}} \rangle   \langle n_{\bm{r}} \rangle   \notag
\end{equation}
where $r$ is the distance between the links $\bm{0}$ and $\bm{r}$. In \figref{fig:spatial_dmr_cor}, $C_d$ between parallel links is plotted as a function of separation, for $\bm{r}$ perpendicular to the orientation of the dimers for a $L=64$ lattice. We find that $C_d$ decays exponentially with distance having a correlation length  $\chi\approx 1$. Additionally, $C_d$ displays a sub-lattice oscillation that is consistent with the finite-size maximum in $S_d$ at $(0,\pi)$. Thus we find that the RK point of the QDPM is a symmetric dimer liquid with exponentially decaying dimer correlations, in contrast with the power-law decaying correlations of the QDM.

@@

\section{Dimer order parameters QDPM RK ground state}
\label{sec:dimer_order_parameters}
%
